---
title: "Multivariate analyses of biodiversity across time and space"
subtitle: "2nd Dynacom Workshop"
author: "Patrick Weigelt"
date: "November 2019"
output: 
  html_document: 
    theme: flatly
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* After having finished "2. Preparation" you can start the tutorial either 3. or 5.*


## 1. Overview  

The main objective for this workshop is to strengthen your biodiversity data analysis skills in R. This tutorial is, however, NOT a cookbook about how to do certain analyses but rather a showcase of two things:

1. There are basic R functions and specialized functions from various R packages that allow you
to reshape and combine data of various kinds (communities, site characteristics, traits, phylogenies) for analyses.

2. There are a lot of dedicated R packages to calculate ecological metrics and do specific community ecological analyses (which we can only touch superficially here).

We use data on palm communities, functional traits, site characteristics, and phylogenetic rekationships as an example dataset due to its unique coverage of different aspects of biodiversity:  


#### Palm communities (plot data):
* Balslev, H., Kristiansen, S.M. & Muscarella, R. (2019). _[Palm community transects and soil properties in western Amazonia.](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2841)_ *Ecology*, e02841.

#### Palm traits:
* Kissling, W.D., Balslev, H., Baker, W.J., Dransfield, J., Göldel, B., Lim, J.Y., Onstein, R.E. & Svenning, J.-C. (2019). _[PalmTraits 1.0, a species-level functional trait database of palms worldwide.](https://www.nature.com/articles/s41597-019-0189-0)_ *Scientific Data*, 6, 178.

#### Palm phylogeny:
* Faurby, S., Eiserhardt, W.L., Baker, W.J. & Svenning, J.-C. (2016). _[An all-evidence species-level supertree for the palms (Arecaceae).](https://www.sciencedirect.com/science/article/pii/S1055790316000701)_ *Molecular Phylogenetics and Evolution*, 100, 57-69.


#### Further reading
Great reading on programming, data analysis and statistics in R can, for example, be found here:

* Wickham, H. (2019). _[Advanced R.](https://adv-r.hadley.nz/)_  
* Wickham, H. (2017). _[R for Data Science.](https://r4ds.had.co.nz/)_  
* Bolker, B. (2008). _[Ecological Models and Data in R.](https://www.degruyter.com/viewbooktoc/product/540514)_  
* Lefcheck, J. (2019). _[Structural Equation Modeling in R for Ecology and Evolution.](https://jslefche.github.io/sem_book/)_  
* Zuur, A., Ieno, E.N., Walker, N., Savaliev, A.A. & Smith, G.M. (2009). _[Mixed Effects Models and Extensions in Ecology with R.](https://www.springer.com/de/book/9780387874579)_  
* Borcard, D., Gillet, F., Legendre, P. (2018). _[Numerical Ecology with R.](https://www.springer.com/de/book/9783319714035)_  
* Lovelace, R., Nowosad, J. & Muenchow, J. (2019). _[Geocomputation with R.](https://geocompr.robinlovelace.net/)_  


## 2. Preparation  

#### a) Install software  

* Get R from <http://www.r-project.org/>
* Get RStudio from <https://www.rstudio.com/>

#### b) Create a new project in RStudio  

* Open RStudio  
* File | New Project  
* Select *either* 'New Directory' or 'Existing Directory', choose a location and name your project  
  (this will be the default working directory for today)  
* Within your project's folder, create three sub-folders:  
    + data  
    + code  
* Create a new empty R script: File | New File | R script  
* Save file in ~/code 
* Start coding!  


#### c) Data  

* Download, unzip and copy data into ~/code  
1. _[Community data](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2841)_  
2. _[Traits and phylogeny](https://datadryad.org/stash/dataset/doi:10.5061/dryad.ts45225)_  


#### d) R packages

* load all nessecary libraries (install the missing ones using `install.packages()`):  

```{r, include=FALSE}
library(maps)
library(plyr)
library(dplyr)
library(ggplot2)
library(psych)
library(vegan)
library(FactoMineR)
library(missMDA)
library(iNEXT)
library(reshape2)
library(FD)
library(sjPlot)
library(jtools)
library(ape)
library(PhyloMeasures)
library(tidyr)
library(Hmisc)
library(car)
library(RColorBrewer)
library(picante)
library(gridExtra)
```

```{r, eval=FALSE}
library(maps)
library(plyr)
library(dplyr)
library(ggplot2)
library(psych)
library(vegan)
library(FactoMineR)
library(missMDA)
library(iNEXT)
library(reshape2)
library(FD)
library(sjPlot)
library(jtools)
library(ape)
library(PhyloMeasures)
library(tidyr)
library(Hmisc)
library(car)
library(RColorBrewer)
library(picante)
library(gridExtra)
```


## 3. Getting started: Plot data  

### a) Palm plots  

#### Clean workspace  
```{r, echo=TRUE}
rm(list=ls())

```

#### Load plot information  
```{r, echo=TRUE}
plots <- read.csv("data/WAmazonian_Palms_transect.csv")
str(plots[,c(1:10)])
names(plots)
range(plots$Latitude)
range(plots$Longitude)
```


#### Map plots  
```{r, echo=TRUE,fig.align='center'}
map('world', xlim = c(-85,-47), ylim = c(-21,13), lforce="e")
title(main = "Western Amazonian palm plots", xlab="longitude", ylab="Latitude")
points(plots$Longitude, plots$Latitude, pch=4)
map.axes()
```

### b) Soil data

#### Load soil data for plots and subplots
```{r, echo=TRUE}
soil <- read.csv("data/WAmazonian_Palms_soil.csv")
tail(soil)
```


#### Summarize soil data at the level of plots

* base R version: 
Apply a function to groups of values (here plots)
```{r, echo=TRUE}
moisture <- tapply(soil$moist, soil$Transect, FUN = mean)
length(which(!is.na(moisture) & !is.nan(moisture) & is.finite(moisture)))
moisture <- tapply(soil$moist, soil$Transect, FUN = function(x) mean(x, na.rm=TRUE))
length(which(!is.na(moisture) & !is.nan(moisture) & is.finite(moisture)))
moisture <- tapply(soil$moist, soil$Transect, FUN = mean, na.rm=TRUE)
length(which(!is.na(moisture) & !is.nan(moisture) & is.finite(moisture)))
moisture[10:30]
```


* base R version: 
Apply a function to groups of values across a whole data.frame
```{r, include=FALSE}
suppressWarnings(soils_plots <- aggregate(soil, list(soil$Transect), FUN = function(x) mean(x, na.rm=TRUE)))
```
```{r, eval=FALSE}
soils_plots <- aggregate(soil, list(soil$Transect), FUN = function(x) mean(x, na.rm=TRUE))
head(soils_plots)
```
The function `mean()` can not be applied to the character columns: NA's introduced.  
We can instead apply different function for different parts of the data.frame but this is a bit tedious:  
```{r, echo=TRUE}
soils_plots_1 <- aggregate(soil[,c(3:29)], list(soil$Transect), FUN = function(x) mean(x, na.rm=TRUE))
soils_plots_2 <- aggregate(soil[,c(30:ncol(soil))], list(soil$Transect), FUN = function(x) paste(unique(x[!is.na(x)]), collapse=","))

soils_plots <- join(soils_plots_1,soils_plots_2, by="Group.1")

head(soils_plots)
```

* *dplyr* version: 
Many packages help with data manipulation (plyr, reshape2, data.table, purrrlyr)  
Most widely used nowadays: _[dplyr](https://dplyr.tidyverse.org/)_
(Also see tutorial from first day by Nathaly Guerrero-Ramirez)

```{r, echo=TRUE}
soil_grouped <- group_by(soil, Transect) ### group by plot
soils_plots <- dplyr::summarize(soil_grouped, mean_moist = mean(moist, na.rm = TRUE)) ### calculate mean values by grouping
```

* *tidy* version: 

```{r, echo=TRUE}
soils_plots <- soil %>%
  group_by(Transect) %>% 
  dplyr::summarize(mean_moist = mean(moist, na.rm = TRUE)) 

```


```{r, echo=TRUE}

soils_plots_1 <- summarize_at(soil_grouped, .vars=c(2:28), .funs = mean, na.rm=TRUE)
soils_plots_2 <- summarize_at(soil_grouped, .vars=c(29:31), .funs = function(x) paste(unique(x[!is.na(x)]), collapse=","))
soils_plots <- join(soils_plots_1,soils_plots_2, by="Transect")

```
#### Aggreating pH  
Of course there are many other options than the arithmetic mean for aggregating data (median, sd, etc). In the case of pH (which is -log10(H+) it would be straightforward to backtransform before calulating the mean. This could be done beforehand on the entire ph columns or within the function for aggregating:

* Beforehand
```{r, echo=TRUE}
soil$H <- 10^(-soil$pH_KCl) # backtransform
soil_grouped <- group_by(soil, Transect) ### group by plot
soils_pH <- dplyr::summarize(soil_grouped, mean_H = mean(H, na.rm = TRUE)) ### calculate mean values by grouping
soils_pH$pH_KCl <- -log10(soils_pH$mean_H) # transform to pH
```

* Within function
```{r, echo=TRUE, fig.width=5,fig.align='center'}
soil_grouped <- group_by(soil, Transect) ### group by plot
soils_pH_2 <- dplyr::summarize(soil_grouped, mean_pH_KCl = -log10(mean(10^-pH_KCl, na.rm = TRUE))) ### apply function to values by grouping

all(soils_pH$pH_KCl==soils_pH_2$mean_pH_KCl, na.rm=TRUE)
plot(soils_pH$pH_KCl, soils_plots$pH_KCl, xlab="mean pH backtransformed", ylab="mean pH")
soils_plots$pH_KCl <- soils_pH$pH_KCl
```
I am not a soil biologist: Think about how to aggregate!  

#### Join main plot data and soil data
```{r, echo=TRUE}
plots <- join(plots, soils_plots, type="left", by="Transect")
#summary(plots)
names(plots)
```




### c) Distribution and collinearity of plot characteristics 


#### Remove plots and variables with too many NAs

```{r, echo=TRUE, fig.width=8,fig.height=8}
plots_PCA <- plots[!(is.na(plots$pH_KCl) & is.na(plots$moist) & is.na(plots$ECEC)),]
plots_PCA <- plots_PCA[,apply(plots_PCA, 2, function(x) sum(is.na(x))<=44) ] 
dim(plots_PCA)
#summary(plots_PCA)
```


#### Plot frequency distributions
```{r, echo=TRUE, fig.width=8, fig.height=8,fig.align='center'}
par(mfrow=c(4,4))
for (i in 1:16){
   hist(plots_PCA[,c(4,7,15:28)][,i], xlab = names(plots_PCA[,c(4,7,15:28)][i]), main="")
}
```

#### Perform log transformation
```{r, echo=TRUE, fig.width=8, fig.height=8,fig.align='center'}
variables_log <- c("Elevation", "moist", "mcf", "loi", "acidity", "ECEC", "Al", "Ca", "Mg", "Na", "K", "P_Bray")

for (i in 1:length(variables_log)){
  plots_PCA[,variables_log[i]] <- log10(plots_PCA[,variables_log[i]]+1)
}

Hmisc::hist.data.frame(plots_PCA[,c(4,7,15:28)])
```

#### Collinearity
```{r, echo=TRUE, fig.width=8, fig.height=8,fig.align='center'}
#pairs.panels(plots_PCA[,unlist(lapply(plots_PCA, is.numeric))])
pairs.panels(plots_PCA[,c(4,7,15:20)])
pairs.panels(plots_PCA[,c(21:28)])
```


### d) Principal Component Analysis (PCA)  

Principal Component Analysis (PCA) is a tool to reduce dimensionality in a dataset when we can expect linear relationships between variables.
```{r, echo=TRUE}
pca<-PCA(plots_PCA[,c(4,7,15:28)],scale.unit = TRUE, graph=FALSE)
```

#### Impute missing values  
https://youtu.be/YDbx2pk9xNY  

* Estimate the number of dimensions for the PCA by cross-validation  
```{r, echo=TRUE, results="hide"}
number_dim <- estim_ncpPCA(plots_PCA[,c(4,7,15:28)], ncp.min=3, ncp.max=5, scale=TRUE, method.cv="kfold", nbsim = 10) # Put nbsim up (to 100); takes longer!
```
```{r, echo=TRUE}
number_dim
```

* Impute missing values  
```{r, echo=TRUE}
plots_PCA_imputed<-imputePCA(plots_PCA[,c(4,7,15:28)], ncp=number_dim$ncp)
plots_PCA_imputed<-as.data.frame(plots_PCA_imputed$completeObs)
rownames(plots_PCA_imputed) <- plots_PCA$Transect

head(plots_PCA[is.na(plots_PCA$pH_KCl),c(4,7,15,16)])
head(plots_PCA_imputed[is.na(plots_PCA$pH_KCl),c(1:4)])
```

#### Run PCA on imputed dataset  
```{r, echo=TRUE,fig.align='center'}
pca_imputed<-PCA(plots_PCA_imputed, scale.unit=TRUE, graph=FALSE) # 60% of variation on first two axes
plot(pca_imputed,choix="var")
# biplot(pca_imputed$ind$coord[,c(1:2)], pca_imputed$var$coord[,c(1:2)])
```


#### Extract axis scores and eigenvalues
```{r, echo=TRUE}
#str(pca_imputed)
head(pca_imputed$eig)
head(pca_imputed$ind$coord)

```


#### Add PCA axis scores to plots_PCA data.frame
```{r, echo=TRUE}
plots_PCA <- cbind(plots_PCA, pca_imputed$ind$coord)
```


## 4. Species diversity and composition

### a) Species composition data

#### Load species occurrances per plot
```{r, echo=TRUE}
specsbysites_long <- read.csv("data/WAmazonian_Palms_species.csv", stringsAsFactors = FALSE)
head(specsbysites_long)
```

#### Simplify data  
Get rid of transect subunits and infraspecific taxon names.
```{r, echo=TRUE}
specsbysites_long <- specsbysites_long[,c("Transect", "Genus", "Species", "Growth_class")]
```

#### Species-by-sites matrix
```{r, echo=TRUE}
specsbysites_long$Species_full <- paste(specsbysites_long$Genus, specsbysites_long$Species, sep=" ")

specsbysites <- table(unique(specsbysites_long[,c("Species_full","Transect")]))
specsbysites <- as.data.frame.matrix(specsbysites)

specsbysites_ab <- table(specsbysites_long[,c("Species_full","Transect")])
specsbysites_ab <- as.data.frame.matrix(specsbysites_ab)
specsbysites_ab[1:5,1:5]
```

#### Species numbers per plot
```{r, echo=TRUE}
colSums(specsbysites)[1:10]
```

#### Individuals per plot
```{r, echo=TRUE}
colSums(specsbysites_ab)[1:10]
```

#### Plots occupied per species
```{r, echo=TRUE}
rowSums(specsbysites)[1:10]
```

#### Individuals per species
```{r, echo=TRUE,fig.align='center'}
rowSums(specsbysites_ab)[1:10]

par(mfrow=c(1,2))
hist(colSums(specsbysites), xlab = "Species number", main="")
hist(colSums(specsbysites_ab), xlab = "Number of individuals", main="")
```


### b) Diversity: Hill numbers  

iNext provides three measures of Hill numbers of order q:  
1. species richness (q = 0)  
2. Shannon diversity (q = 1, the exponential of Shannon entropy)  
3. Simpson diversity (q = 2, the inverse of Simpson concentration)  
For each diversity measure, iNEXT uses the observed sample of abundance or incidence data to compute diversity estimates for two types of rarefaction and extrapolation.  

Learn more here:  
https://cran.r-project.org/web/packages/iNEXT/vignettes/Introduction.html  

Read about effective numbers of species and how to get them based on diversity indices like Shannon and Simpson:  
http://www.loujost.com/Statistics%20and%20Physics/Diversity%20and%20Similarity/EffectiveNumberOfSpecies.htm  


#### Calculate diversity metrics  
```{r, echo=TRUE}
hill_numbers <- iNEXT(specsbysites_ab, nboot = 10) # put up nboot (to 50), takes longer
```


```{r, echo=TRUE}
# head(hill_numbers$DataInfo) # basic data information
# head(hill_numbers$$iNextEst) # Diversity estimates with rarefied and extrapolated samples.
```

Plotting accumulation curves would make sense for groups of plots for example from a certain region or treatment. Here we only use the hill numbers and diversity estimates per plot:
```{r, echo=TRUE}
head(hill_numbers$AsyEst) # asymptotic diversity estimates according to Chao et al. (2013)
```

#### Join diversity metrics to plot data

To join the diversity metrics to our plots we need to convert them to wide format. Currently semi-long format?  

* Since empty spaces in column names are unhandy we first replace them in the Diversity column
```{r, echo=TRUE}
hill_numbers$AsyEst$Diversity <- gsub(" ", "_", hill_numbers$AsyEst$Diversity)
head(hill_numbers$AsyEst)
```

* Reshape long format data.frame to join to plots using reshape2
```{r, echo=TRUE}
diversity <- melt(hill_numbers$AsyEst, id.vars=1:2)
head(diversity)
diversity <- dcast(diversity, Site~variable+Diversity, value.var = "value")
head(diversity)
```

* Reshape long format data.frame to join to plots using tidyr
```{r, echo=TRUE}
diversity_2 <- pivot_wider(hill_numbers$AsyEst, id_cols=c(Site,Diversity), names_from = Diversity, values_from=c(Observed,Estimator,s.e.,LCL,UCL))
head(diversity_2)
```

* Join plots and diversity
```{r, echo=TRUE,fig.align='center', fig.height=8, fig.width=8}
names(diversity)[1] <- "Transect"
plots_PCA <- join(diversity, plots_PCA, type="right", by="Transect")
pairs.panels(diversity[,c(2:7)])
```

### c) Diversity - environment relationship
```{r, echo=TRUE,fig.align='center'}
par(mfrow = c(2,3))
plot(abs(plots_PCA$Latitude), plots_PCA$Observed_Species_richness, xlab = "abs Latitude", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~abs(plots_PCA$Latitude)), col="darkred")
plot(plots_PCA$Mean_annual_PPT, plots_PCA$Observed_Species_richness, xlab = "Precipitation", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~plots_PCA$Mean_annual_PPT), col="darkred")
plot(plots_PCA$Elevation, plots_PCA$Observed_Species_richness, xlab = "Elevation", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~plots_PCA$Elevation), col="darkred")

plot(plots_PCA$Dim.1, plots_PCA$Observed_Species_richness, xlab = "PCA axis 1", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~plots_PCA$Dim.1), col="darkred")
plot(plots_PCA$Dim.2, plots_PCA$Observed_Species_richness, xlab = "PCA axis 2", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~plots_PCA$Dim.2), col="darkred")
plot(plots_PCA$Dim.3, plots_PCA$Observed_Species_richness, xlab = "PCA axis 3", ylab = "Species richness", pch=16, col=rgb(0,0,0,0.25), cex=0.5)
regLine(lm(plots_PCA$Observed_Species_richness~plots_PCA$Dim.3), col="darkred")
```


#### Linear Multi-predictor model
```{r, echo=TRUE,fig.align='center', fig.width=8}
richness_model <- lm(Observed_Species_richness ~ Habitat_type + Dim.1 + Dim.2 + Dim.3, data = plots_PCA)
summary(richness_model)
#crPlots(richness_model, main = "", ylab="Species Richness", smooth=FALSE, grid=FALSE)

p1 <- effect_plot(richness_model, pred = "Habitat_type", interval = TRUE, x.label="Species Richness", y.label="PCA axis 1", point.size = 2, partial.residuals = TRUE, jitter = .2)
p2 <- effect_plot(richness_model, pred = "Dim.1", interval = TRUE, partial.residuals = TRUE,      x.label="Species Richness", y.label="PCA axis 1", point.size = 2)
p3 <- effect_plot(richness_model, pred = "Dim.2", interval = TRUE, partial.residuals = TRUE,      x.label="Species Richness", y.label="PCA axis 2", point.size = 2)
p4 <- effect_plot(richness_model, pred = "Dim.3", interval = TRUE, partial.residuals = TRUE,      x.label="Species Richness", y.label="PCA axis 3", point.size = 2)
grid.arrange(p1, p2, p3, p4, nrow = 2)
```
We are not going into the details of regression models and model selection here.  
For your experimental setup you will most likely use mixed effects models. New package for mixed model diagnostics: https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html  
For plotting: https://cran.r-project.org/web/packages/jtools/vignettes/effect_plot.html  

### d) Non-metric multidimensional scaling (NMDS)

#### Calculate species turnover

* Subset our species by sites matrix for plots included in plots_PCA
```{r, echo=TRUE}
specsbysites_ab_PCA <- specsbysites_ab[,which(names(specsbysites_ab) %in% plots_PCA$Transect)]
specsbysites_ab_PCA <- specsbysites_ab_PCA[which(rowSums(specsbysites_ab_PCA)>0),]
```
  
* Order plots the same way like species by sites
```{r, echo=TRUE}
all(row.names(plots_PCA_imputed)==names(specsbysites_ab_PCA))
plots_PCA_imputed <- plots_PCA_imputed[order(row.names(plots_PCA_imputed)),]
plots_PCA <- plots_PCA[order(plots_PCA$Transect),]
all(plots_PCA$Transect==names(specsbysites_ab_PCA))
```
  
* Calculate Bray Curtis dissimilarity, geographic and environmental distances
```{r, echo=TRUE}
betadiv <- vegdist(t(specsbysites_ab_PCA), method="bray")  # 1- abundance based Sörensen Index
geodist <- vegdist(plots_PCA[,c("Latitude", "Longitude")], method="euclidean")
envdist <- vegdist(scale(plots_PCA_imputed), method="euclidean")
```
  
* Plot them
```{r, echo=TRUE,fig.align='center'}
par(mfrow=c(1,2))
plot(geodist, betadiv, xlab="Geographic distance", ylab = "Palm turnover", pch=16, col=rgb(0,0,0,0.25), cex=0.2)
plot(envdist, betadiv, xlab="Environmental distance", ylab = "Palm turnover", pch=16, col=rgb(0,0,0,0.25), cex=0.2)
```
  
Have a look at Generalized dissimilarity modelling:  
Ferrier et al. (2007). _[Using generalized dissimilarity modelling to analyse and predict patterns of beta diversity in regional biodiversity assessment.](https://onlinelibrary.wiley.com/doi/10.1111/j.1472-4642.2007.00341.x)_ *Diversity and Distributions*, 17, 252-264.  


#### Run NMDS  
NMDS tries to reduce the difference between distances in the original matrix (here turnover) and distances in the ordination. The difference is quantified as stress. Lower stress value = better ordination. Clarke (1993): <0.05 = excellent, <0.10 = good, <0.20 = usable, >0.20 = not acceptable.
```{r, echo=TRUE}
set.seed(1000)
nmds <- metaMDS(comm=t(specsbysites_ab_PCA), k=3, trace=FALSE)
nmds
```
```{r, echo=TRUE,fig.align='center'}
plot(nmds, type="none")
colours <- data.frame(Habitat_type = unique(plots_PCA$Habitat_type), colour=brewer.pal(7,"Dark2"), stringsAsFactors = FALSE)
plots_PCA <- join(plots_PCA, colours, type="left", by="Habitat_type")
points(nmds, display="sites", col=plots_PCA$colour)
legend(-3.5,2,colours$Habitat_type, colours$colour, bty="n")
```

#### Fit environmental vectors
```{r, echo=TRUE,fig.align='center'}
plot(nmds)
ef <- envfit(nmds, plots_PCA_imputed, permu = 999)
plot(nmds)
plot(ef, p.max = 0.05)
```

Check _[vegantutor](http://cc.oulu.fi/~jarioksa/opetus/metodi/vegantutor.pdf)_ for more options


## 5. Species traits

### a) Trait data

#### Load traits  
```{r, echo=TRUE}
traits <- read.table("data/PalmTraits_1.0.txt", stringsAsFactors = FALSE, sep= "\t", header = TRUE)
# View traits
```

#### Reload Species by sites  
We reload the species by sites data in case somebody got lost. Everyone should be able to continue from here now  
```{r, echo=TRUE}
specsbysites_long <- read.csv("data/WAmazonian_Palms_species.csv", stringsAsFactors = FALSE)
```

#### Match trait and distribution data  
* Extract unique species names and create a combined species name (Genus epithet)  
```{r, echo=TRUE}
species <- unique(specsbysites_long[,c("Genus", "Species")])
species$Species_full <- paste(species$Genus, species$Species, sep=" ")
nrow(species) # number of species
sum(species$Species!="sp") # number of identifies species
```

* Check for species missing in traits data 
```{r, echo=TRUE}
species$Species_full[which(!species$Species_full %in% traits$SpecName & species$Species != "sp")]
```
"Aiphanes truncata" is a synonym of Aiphanes horrida according to theplantlist.org  
"Geonoma longepedunculata" is spelled "Geonoma longipedunculata" in the traits table and this is correct according to theplantlist.org  

* Change names
```{r, echo=TRUE}
specsbysites_long$Species[which(specsbysites_long$Genus == "Aiphanes" & specsbysites_long$Species == "truncata")] <- "horrida"
specsbysites_long$Species[which(specsbysites_long$Genus == "Geonoma" & specsbysites_long$Species == "longepedunculata")] <- "longipedunculata"
```

* Get new unique species names and compare again 
```{r, echo=TRUE}
species <- unique(specsbysites_long[,c("Genus", "Species")])
species$Species_full <- paste(species$Genus, species$Species, sep=" ")
length(which(species$Species != "sp"))

species$Species_full[which(!species$Species_full %in% traits$SpecName & species$Species != "sp")]
```
No missing species left!  

* Prepare new species by sites matrices
```{r, echo=TRUE}
specsbysites_long$Species_full <- paste(specsbysites_long$Genus, specsbysites_long$Species, sep=" ")

specsbysites_ab <- table(specsbysites_long[which(specsbysites_long$Species != "sp"),c("Transect", "Species_full")])
specsbysites_ab <- as.data.frame.matrix(specsbysites_ab)
```


* Subset traits
```{r, echo=TRUE}
traits <- traits[which(traits$SpecName %in% species$Species_full),]
summary(traits)
```


### b) Functional diversity

Distance-based approaches use functional pairwise distance matrices calculated with principal components or coordinates analysis on multiple traits. The _[FD package](http://www.imsbio.co.jp/RGM/R_rdfile?f=FD/man/FD-package.Rd&d=R_CC)_ calculates three facets of functional diversity: functional richness, evenness, and dispersion. These facets are qualitatively similar to taxonomic diversity indices, i.e. species richness, evenness, and diversity. See _[Chao et al. (2019)](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1343)_ for latest developments.  


#### Make subset for a couple of interesting traits
```{r, echo=TRUE}
rownames(traits) <- traits$SpecName
traits_FD <- traits[,c("MaxStemHeight_m", "MaxStemDia_cm", "MaxLeafNumber", "Max_Blade_Length_m", "Max_Rachis_Length_m", "AverageFruitLength_cm", "AverageFruitWidth_cm")]
```



#### Look at frequency distributions and transform
```{r, echo=TRUE, fig.width=8, fig.height=8,fig.align='center'}
par(mfrow=c(3,3))

for(i in 1:ncol(traits_FD)){
  hist(traits_FD[,i], main="", xlab=names(traits_FD)[i])
}

```

* Log-transform
```{r, echo=TRUE, fig.width=8, fig.height=8,fig.align='center'}
traits_FD <- data.frame(apply(traits_FD, 2, function(x) log10(x+1)))

par(mfrow=c(3,3))
for(i in 1:ncol(traits_FD)){
  hist(traits_FD[,i], main="", xlab=names(traits_FD)[i])
}
```

#### Calculate Functional diversity metrics
```{r, echo=TRUE}
palms_FD<-dbFD(traits_FD, specsbysites_ab, corr="cailliez", stand.x=TRUE, calc.FRic = TRUE,  
                  stand.FRic = FALSE,calc.FGR=FALSE,calc.CWM=TRUE, calc.FDiv=FALSE)
str(palms_FD)
```
```{r, echo=TRUE, fig.height=4,fig.align='center'}

palms_FD_all <- data.frame(Transect=names(palms_FD$nbsp), nbsp=palms_FD$nbsp, FRic=palms_FD$FRic, FEve=palms_FD$FEve, FDis=palms_FD$FDis)

par(mfrow=c(1,3))
plot(palms_FD_all$nbsp, palms_FD_all$FRic)
plot(palms_FD_all$nbsp, palms_FD_all$FEve)
plot(palms_FD_all$nbsp, palms_FD_all$FDis)
```



### c) Community weighted means
```{r, echo=TRUE, fig.width=6,fig.align='center'}
palms_CWMs <- palms_FD$CWM
palms_CWMs$Transect <- row.names(palms_CWMs)
```


#### Join FD and CMWs to plots
```{r, echo=TRUE, fig.width=6,fig.align='center'}
plots <- read.csv("data/WAmazonian_Palms_transect.csv")
plots <- join(plots, palms_FD_all, by="Transect", type="left")
plots <- join(plots, palms_CWMs, by="Transect", type="left")

plot(plots$Mean_annual_PPT, plots$AverageFruitWidth_cm, xlab="Precipitation", ylab="Mean Fruit Width") 
```



## 6. Phylogeny 

### a) Phylogenetic tree

#### Load tree and plot
```{r, echo=TRUE}
palm_tree <- read.nexus("data/TREE.nex")
str(palm_tree)
```


#### Join tree to palm communities
```{r, echo=TRUE, fig.height=8,fig.align='center'}
names(specsbysites_ab) <- gsub(" ", "_", names(specsbysites_ab))
palm_tree_pruned <- prune.sample(specsbysites_ab, palm_tree)
plot(palm_tree_pruned, type="fan", show.tip.label = TRUE, cex=0.5)
```


### b) Phylogenetic diversity

#### Calculate PD
```{r, echo=TRUE, warning=FALSE, message=FALSE}
palms_PD <- pd.query(palm_tree_pruned, specsbysites_ab, standardize = FALSE)
palms_PDses <- pd.query(palm_tree_pruned, specsbysites_ab, standardize = TRUE,null.model="uniform")
palms_PD <- data.frame(Transect=row.names(specsbysites_ab), palms_PD, palms_PDses)
head(palms_PD)
```

#### Join PD to plots and compare to taxonomic and functional richness
```{r, echo=TRUE,fig.align='center', fig.height=4}
plots <- join(plots, palms_PD, by="Transect", type="left")

par(mfrow=c(1,3))
plot(plots$nbsp, plots$palms_PD, xlab="Species richness", ylab="Faith's PD")
PD_model <- lm(palms_PD ~ nbsp + I(nbsp^2), data=plots)
prediction <- predict(PD_model, newdata = data.frame(nbsp=seq(min(plots$nbsp), max(plots$nbsp), length.out=1000)))
points(seq(min(plots$nbsp), max(plots$nbsp), length.out=1000), prediction, type="l", col="darkred", lwd=2)
plot(plots$nbsp, plots$palms_PDses, xlab="Species richness", ylab="Faith's PDses") 
plot(plots$palms_PD, plots$FRic, xlab="Faith's PD", ylab="Functional Richness")
regLine(lm(plots$FRic~plots$palms_PD), col="darkred", lwd=2)
```